---
title: Tree diversity changes estimation issues. Commentary on Fadrique et al. (2026)
format:
  stylisharticle-html: default
  nature-pdf:
    journal:
      cite-style: sn-nature
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    classoption: [lineno]
author:
  - name: Eric Marcon
    affiliations:
      - name: AgroParisTech
        id: 1
        city: Montpellier
        state: France
      - name: AMAP, CIRAD, CNRS, INRAE, IRD, Univ Montpellier
        id: 2
        city: Montpellier
        state: France
    email: eric.marcon@agroparistech.fr
    attributes:
      corresponding: true
abstract: |
  Issues in the estimation of yearly richness change the Amazon derived by Fadrique *et al.* (2026) in Nature Ecology & Evolution are addressed. Richness change was not standardized to a consistent plot size, preventing comparing it between plots, and the yearly richness change was calculated with a linear approximation leading to its underestimation in long-term plots. The necessary methods are proposed to correct these issues.
  
bibliography: bibliography.bib
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: true
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitr", "formatR"))

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c("tidyverse", "divent", "ggpubr")
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = FALSE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
# Function to call in each chunk
ggtheme_set <- function() {
  theme_set(theme_bw())
  theme_update(
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  )
}
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

# Introduction

Fadrique and colleagues [@Fadrique2026] assess the long-term variations of species richness in the Amazon and explore their determinants.
The way they estimate the yearly richness change in each of the 406 plots used in the study faces two methodological issues that are serious enough to question the results and justify recomputing it.

The consequences of unequal plot sizes are shown in the first section of this reply, followed by those of the linear approximation of richness change along years.
The last section suggests the methods to deal with these issues.

Examples below were made with the *divent* package [@divent] for R [@R].

# Plot sizes

Richness change is characterized in the article as a relative increase or decrease compared to the initial number of species.
Yet, the number of species and its change depend on the area of each plot.
Plots are samples of local communities of trees: their richness depend on the species distribution in the community they belong to and of their size.
The species accumulation curves [@Gotelli2001], i.e. the expected richness with respect to sample size, can be computed easily [@Chao2014].

[@fig-plot_size]a shows the species accumulation curves of two communities.
The initial one is simulated in a log-normal distribution with standard deviation 1.3 and 300 species.
The final one is drawn in the same distribution with 20% less species.
The observed richness change relies heavily on the plot size ([@fig-plot_size]b).
The same changes in the community distribution imply completely different observed changes in plots of different sizes, that may even be contradictory: richness gain here according to a 125-tree (around 0.25-ha) plot versus richness loss in a 500-tree (1-ha) plot.
@fig-plot_size is just an illustration.
Various simulations show unpredictable shapes of the relative richness change curve: the difference of richness clearly increases with plot size but both initial and final richness increase, so their ratio may vary in different directions, even though it will ultimately reach an asymptote (-20% in [@fig-plot_size]b).

::: {#fig-plot_size}
```{r}
#| label: plot_size
ggtheme_set()
library("divent")
the_size <- 1E6
the_richness <- 300
the_sd_lnorm <- 1.3
# Create a large community ("initial")
rcommunity(
  1,
  size = the_size, 
  species_number = the_richness, 
  distribution = "lnorm",
  sd_lnorm = the_sd_lnorm
) %>% 
  mutate(site = "original") %>% 
  # Save it
  {. ->> the_community_300} %>% 
  # Compute SAC
  accum_hill(
    q = 0,
    levels = 1:1000
  ) ->> p6_accum
# Species loss
the_loss <- 20/100
# Create the community ("final")
rcommunity(
  1,
  size = the_size, 
  species_number = the_richness * (1 - the_loss), 
  distribution = "lnorm",
  sd_lnorm = the_sd_lnorm
) %>% 
  {. ->> the_community_240} %>% 
  # Compute SAC
  accum_hill(
    q = 0,
    levels = 1:1000
  ) %>% 
  mutate(site = "loss") %>% 
  # Add the initial community richness
  bind_rows(p6_accum) %>%
  {. ->> the_accum} %>% 
  autoplot() +
    theme(legend.position = "none") ->
  fig_SAC
# Richness change
the_accum %>% 
  pivot_wider(names_from = site, values_from = "diversity") %>% 
  mutate(change = (loss - original) / original) %>% 
  ggplot(aes(x = level, y = change)) +
    geom_line() +
    labs(x = "Sample Size", y = "Relative Richness Change") ->
  fig_change

library(ggpubr)
ggarrange(fig_SAC, fig_change, labels = letters[1:2], ncol = 2)
```

\(a\) Species accumulation curves of a log-normal community with 300 species (blue) or 240 species (red).
(b) The observed relative richness change depends on the sample size.
:::

Richness change thus cannot be compared between plot of different sizes: the same variation in the species distribution of the local community has quantitatively and, for small plots, sometimes qualitatively different effects on richness change.

# Yearly richness change

```{r}
#| label: r_text
# r calculated by linear approximation. Real value = 1%
r_linear_1pc <- format((150 - 150 * .99^40) / 150 / 40 * 100, digits = 2)
# Real value = 5%
r_linear_5pc <- format((150 - 150 * .95^40) / 150 / 40 * 100, digits = 2)
```

::: {#fig-species_loss}
```{r}
#| label: species_loss
ggtheme_set()
# Initial number of species
sp_initial <- 150
# Yearly change rate
r <- -1/100
# Time
years <- 0:40
# Number of species along time
sp_year <- sp_initial * (1 + r)^years
# Linear approximation
sp_linear <- sp_initial + (min(sp_year) - sp_initial) / max(years) * years

# Plot
tibble(time = years, exponential = sp_year, linear = sp_linear) %>% 
  pivot_longer(cols = -"time", names_to = "model") %>% 
  ggplot(aes(x = time, y = value, color = model)) +
    geom_line() +
    labs(x = "Time", y = "Richness") +
    theme(legend.position = "none")
```

1% yearly richness loss along 40 years.
The number of species is represented by the red curve, from 150 species initially to approximately 100 species after 40 years.
The blue line is the incorrect linear approximation of the curve based on the initial and final numbers of species, that underestimates the yearly richness loss: its slope is -`r r_linear_1pc`% of the initial number of species.
:::

In the article, yearly richness change is calculated as

$$
r = ({SP}_{final} - {SP}_{initial}) / {SP}_{initial} / {time}
$$ {#eq-linear_r}

even though it is incorrectly reported as the opposite in the text.
The multiplication by 100 to obtain a percentage is omitted here.
This equation can be rearranged as

$$
{SP}_{final} = {SP}_{initial} * (1 + r * {time}).
$$ {#eq-linear}

The purpose is to estimate an average yearly richness change, whatever the time interval.
Then, richness at time $t+1$ is richness at time $t$ multiplied by $1+r$, whatever $t$ during the interval.
Thus, the correct way to calculate $r$ is derived from the relation

$$
{SP}_{final} = {SP}_{initial} * (1 + r)^{time}.
$$ {#eq-exponential}

This is the discrete form of an exponential variation.
The linear approximation (@eq-linear) used in the article is acceptable for small intervals but leads to large error over decades.

The correct richness loss is obtained by solving the previous equation:

$$
r = \left({SP}_{final} / {SP}_{initial}\right)^{1 / {time}} - 1.
$$ {#eq-exponential_r}

An example is shown in @fig-species_loss.
Starting from 150 species, $r$ is arbitrarily chosen equal to 1% per year.
After 40 years, the final number of species (@eq-exponential) is close to `r round(150 * .99^40)`, rather than 90 according to @eq-linear.
@eq-linear_r used in the article estimates the species loss to be `r r_linear_1pc`% instead of 1%.

The error increases with the absolute value of richness change: 5% yearly species lost during 40 years is estimated to be only `r r_linear_5pc`% by @eq-linear_r.
Richness change absolute value is thus systematically underestimated in the article.
The error increases with the interval between inventories and its actual value, making the more valuable data less reliable.

# Appropriate methods

::: {#fig-standardized}
```{r}
#| label: standardized
ggtheme_set()
simulations_n <- 100
# Simulate plots. Choose plot size and species distribution
rcommunity(
  simulations_n, 
  size = 1500, 
  abd = as.numeric(the_community_300)
) %>% 
  # Compute standardized richness
  div_richness(level = 500) %>% 
  # Remember the name of the community
  mutate(plot = "Initial")  %>% 
  # Bind richness values of other simulations
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 1250, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 1000, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 750, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 500, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 250, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 125, 
      abd = as.numeric(the_community_300)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Initial")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 1500, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final") 
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 1250, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final") 
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 1000, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final") 
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 750, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final") 
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 500, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final") 
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 250, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final")
  ) %>% 
  bind_rows(
    rcommunity(
      simulations_n, 
      size = 125, 
      abd = as.numeric(the_community_240)
    ) %>% 
      div_richness(level = 500) %>% 
      mutate(plot = "Final")
  ) %>%
  # Calculate mean diversity
  group_by(plot, weight) %>% 
  {. ->> richness_sims} %>% 
  summarize(
    richness_mean = mean(diversity),
    richness_sd = sd(diversity)
  ) %>% 
  # Plot
  ggplot(aes(x = weight, y = richness_mean, color = plot)) +
    geom_vline(xintercept = 500, lty = 2) +
    geom_line() +
    geom_point() +
    geom_errorbar(
      aes(
        ymin = richness_mean - richness_sd, 
        ymax = richness_mean + richness_sd
      ),
      width = 40,
      position = position_dodge(10)
    ) +
    labs(x = "Plot Size", y = "Standardized Richness") +
    theme(legend.position = "none") ->
  fig_richness

richness_sims %>% 
  # Calculate richness change
  pivot_wider(
    names_from = plot, 
    values_from = c(diversity)
  ) %>%
  mutate(change = (Final - Initial ) / Initial ) %>% 
  # Calculate the mean richness change
  group_by(weight) %>% 
  summarize(
    change_mean = mean(change), 
    change_sd = sd(change)
  ) %>% 
  ggplot(aes(x = weight, y = change_mean)) +
    geom_vline(xintercept = 500, lty = 2) +
    geom_line() +
    geom_point() +
    geom_errorbar(
      aes(
        ymin = change_mean - change_sd, 
        ymax = change_mean + change_sd
      ),
      width = 40
    ) +
    labs(x = "Plot Size", y = "Relative Richness Change") ->
  fig_change

ggarrange(fig_richness, fig_change, labels = letters[1:2], ncol = 2)
```

\(a\) Richness standardized at 500 trees of samples of different sizes from a log-normal community with 300 species (blue) or 240 species (red).
The lines are the mean values of 100 simulations, and the bars correspond to one standard deviation.
(b) The observed relative richness change does not depend on the sample size when it is at least half the standard level of extrapolation.
:::

These issues can be corrected relatively simply.
The first step consists of standardizing the plot size.
Its choice has consequences on the results and their significance: the smaller the plots, the lower the ability to take into account species that are not the most abundant [@Grabchak2016].
The richness of larger plots can be rarefied with no limit and that of smaller plots extrapolated [@Gotelli2001].

[@fig-standardized]a shows the average richness estimation, standardized at 500 trees (1 ha, assuming density is 500 trees/ha), of 100 simulated plots of sizes varying from 125 to 1500 trees (1/4 to 3 ha) drawn in the communities of 300 and 240 species.
Richness change is calculated for each pair of simulated plots and represented in [@fig-standardized]b.
In contrast with [@fig-plot_size]b, it is independent from plot size.

Extrapolation over twice the original size is considered risky [@Chao2014] (the 125-trees plots in [@fig-standardized]b do not allow estimating richness change correctly) so the standard size may have to be reduced if too many small plots are used.
The trade-off is between larger standardized plots, including information about more species and smaller plots to limit extrapolation uncertainty.
Since the average plot size is 1.23 ha, 1 ha may be a good choice: a round size, used in many studies, has the advantage of allowing intuitive interpretation.

In practice, R packages *iNEXT* [@Hsieh2014] or *divent* (used here) allow interpolating and extrapolating richness of an abundance distribution to the desired number of individuals, that must be derived from each plot's tree density to reach the target area: for instance, extrapolating the diversity of a half-hectare plot with 300 trees up to one hectare area implies extrapolating up to 600 trees.

The standardized initial and final richness of each plot will then be available.
The second step is calculating richness change according to @eq-exponential_r.
Then, all analyzes can be run again.

# References

::: {#refs}
:::
